---
title: "Mokelumne Stream Flows"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: set-up
#| warning: false
#| message: false
#| echo: false

library(tidyverse)
library(dataRetrieval)
```

To pull NWIS data, we need the following information:

-   The type of data (groundwater, unit, water quality, daily, etc..)
-   USGS site number(s)
-   USGS parameter code(s)
-   Time frame (start and end date)

Returned data also has several attributes attached to the data frame. Check with: names(attributes())

| Data type | Name                          | Service                         |
|--------------|-----------------------------|-----------------------------|
| Unit      | Regular frequency (real-time) | uv                              |
| Daily     | Aggregated to daily stats     | dv                              |
| Discrete  | Non-regular time collections  | qw, gwlevel, rating, peak, meas |

| PCode | Name        | Stat Code | Name   |
|-------|-------------|-----------|--------|
| 00060 | Discharge   | 0001      | Max    |
| 00065 | Gage height | 0002      | Min    |
| 00010 | Temp        | 0003      | Mean   |
| 00400 | pH          | 0008      | Median |

```{r}
#| label: pull-data

site_num <- c(
  11315000,  # Cole Creek upper
  11315030,  # Cole Creek below diversion dam
  11314000,  # Tiger Creek below Salt Spring Dam
  11314500,  # NF Mokelume River below Salt Spring Dam
  11315900,  # Bear River below Bear River Dam
  11316100,  # Bear River below Brar River diversion dam
  11316670  # NK Mokelumne below Tiger Creek
)
site_num <- "11314500"
p_code <- "00060"
start_date <- "1900-01-01"
end_date <- "2023-08-01"

moke <- readNWISdv(
  siteNumbers = site_num,
  parameterCd = p_code,
  startDate = start_date,
  endDate = end_date
  )

moke <- renameNWISColumns(moke)
names(moke)
head(moke)

# check attributes
names(attributes(moke))
attr(moke, "variableInfo")
attr(moke, "siteInfo")

# get attributes
parameterInfo <- attr(moke, "variableInfo")
siteInfo <- attr(moke, "siteInfo")

# add station names
moke <- moke |> 
  left_join(siteInfo |> select(site_no, station_nm), by = "site_no") 
```

All stations over time

```{r}
moke |> 
  ggplot(aes(Date, Flow)) + 
  facet_wrap(vars(station_nm), scales = "free") + 
  geom_line() + 
  labs(
    x = "", 
    y = parameterInfo$variableDescription
    )
```

North Fork Mokelumne

```{r}
moke |> 
  filter(site_no %in% c(11314500, 11316670)) |> 
  ggplot(aes(Date, Flow)) + 
  facet_wrap(vars(station_nm), scales = "free") + 
  geom_line() + 
  labs(
    x = "", 
    y = parameterInfo$variableDescription
    )
```


```{r}
moke |> 
  filter(site_no %in% c(11315000, 11315030)) |> 
  ggplot(aes(Date, Flow)) + 
  facet_wrap(vars(station_nm), scales = "free") + 
  geom_line() + 
  labs(
    x = "", 
    y = parameterInfo$variableDescription
    )
```
