---
title: "Literature Review"
format: html
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: prep
#| warning: false
#| message: false
#| echo: false

library(tidyverse)
library(here)
library(kableExtra)
```

This literature review aims to evaluate the streamflow and temperature requirements of aquatic species to assess the significance of potential changes to the stream thermal regime from Mokelumne. We also review thermal criteria set by resource agencies (e.g., NMFS)

## Fish and Aquatic Organisms

The Project's affected area for fish and aquatic resources is located within the Cole Creek, Bear River, Upper and Lower Bear Reservoirs and the Salt Springs Reservoir -- North Fork Mokelumne River watersheds. More specifically, the proposed Project will involve the Salt Springs Reservoir, and the Upper and Lower Bear Reservoirs, all of which are on the Mokelumne River. Additionally, the proposed water conveyance tunnel alignments that will connect the Salt Springs and Bear Reservoirs, pass beneath Cole Creek, a tributary of the Mokelumne.

The Project watersheds feature similar species composition, hydrologic regime, and topographic profiles to other western alpine watersheds in California. These three watersheds are located on the western slope of the Sierra Nevada Mountain Range and are therefore primarily fed by snow melt runoff accumulated from higher elevations.

### Fishes

```{r}
#| label: fish-species
#| eval: false
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Fish species identified by GreenGen in study area."
#| tbl-colwidths: [15,40,40,10]
species <- read_csv(here("data","species-list.csv"))
species |> filter(Class == "Fishes") |> knitr::kable()

```

| Common Name                                | Scientific Name                   | Origin |
|---------------------------|----------------------------|-----------------|
| Pacific Lamprey                            | *Entosphenus tridentatus*         | Native |
| Delta Smelt                                | *Hypomesus transpacificus*        | Native |
| Hardhead                                   | *Mylopharaodon conocephalus*      | Native |
| Sacramento Hitch                           | *Lavinia exilicauda exilicauda*   | Native |
| Chinook Salmon (Central Valley spring run) | *Oncorhynchus tshawytscha* pop. 6 | Native |
| Steelhead (coastal Rainbow Trout)          | *Oncorhynchus mykiss irrideus*    | Native |

: Fish species identified by GreenGen in study area.

### Reptiles and amphibians

| Common Name                      | Scientific Name                        | Origin |
|---------------------------|----------------------------|-----------------|
| Yosemite toad                    | *Anaxyrus canorus*                     | Native |
| Foothill yellow-legged frog      | *Rana boylii*                          | Native |
| Sierra Nevada yellow-legged frog | *Rana sierrae*                         | Native |
| Southern long-toed salamander    | *Ambystoma macrodactylum sigillatum*   | Native |
| Western pond turtle              | *Emys marmorata / Actinemys marmorata* | Native |

: Amphibian and reptile species with potential to occur in study area.

Potential impacts identified in Mokelumne RSP:

-   water quality and quantify (temperature and instream/environmental flows)
-   habitat quality and quantify (availability)
-   movement of pathogens (chytridiomycosis fungus *Batrachochytrium dendrobatidis* and *Ranavirus* spp.)

## Streamflow requirements

## Temperature Requirements

## Flow and Thermal Criteria in Region

## Pumped Hydro Review

### @blakers2021: A review of pumped hydro energy storage

-   Off-river pumped hydro as a very large-scale low-cost and mature energy storage technology to support 100% renewable energy futures
-   Off river hydro is likely to have low environmental impact and low water consumption
    -   Little focus on actual ecological impacts - mostly large scale land/water use
    -   Effectively says no ecological impact on off-river, so long as placed in benign place
-   data from @irena2020, IEA [Data Page](https://www.iea.org/data-and-statistics?country=WORLD&fuel=Energy%20supply&indicator=Electricity%20generation%20by%20source), US DOE [Global ESS Database](https://gesdb.sandia.gov/)
-   impacts primary relate to flooding of land within reservoir
-   other impacts: construction of roads, pipes/tunnels for conveying water, powerhouse, switchyard, high voltage transmission lines
-   How much storage capacity is needed?
    -   rule of thumb = 1 d (24h) storage, accommodates day-night cycle (batteries & PSH), about 27 GWh per million people
-   How much land is needed?
    -   small for off-river and can be selected to minimize sensitive land
    -   per GWh, 12 ha for off-river PHES, head of 400m, 90% efficient, 85% usable water volume, 20 m average water depth = 3 m\^2 per person (queen size bed)
    -   For 27 GWh per 1 million people = 1 plant with 1 GW power and 3 km\^2 flooded area
    -   20x less land needed than for solar PVs
    -   then add floating PVs which can also reduce evaporation
-   How much water is needed?
    -   per GWh, 1 Gigalitre for off-rover PHES, head of 400m, 90% efficient
    -   For 27 GWh per 1 million people = 27 kl per person - NOT CONSUMED!
    -   water needed for initial fill and to replace evaporation is 3 l/person/day, which is about 20 s of a typical shower
    -   Evaporation suppressors can tip the balance in favor of rainfall in arid regions

#### Future energies & big picture

-   Major economies (USA, EU, China, Japan) committed to 100% renewables by 2050s
-   global human population at 10 billion by 2050; when developing counties reach advanced needs, world will need 200,000 TWh/year
-   assuming solar (60%), wind (30%), other (10%), need 81 TW solar, 17 TW wind
-   assuming 1 day rule of thumb, need 500 TWh (storage energy) and 20 TW (power)
    -   more than order of magnitude larger than now, but way less the available pool of off-rover pumped hydro resources (23,000 TWh)

### @stocks2021: global atlas of closed-loop pumped hydro

-   ANU 100% Renewable Energy Group created [global pumped hydro atlas](https://re100.eng.anu.edu.au/index.html)
    -   616,000 off-river sites; 23 million GW-hours = 100x more than needed for 100% electricity world
-   US = 35,000 sites, 1.4 million GWh = 300x more than needed to support solar/wind
-   All reservoirs are outside national parks or urban areas
    -   combined reservoir area is small (3 m\^2 per person per day)
    -   water requirement is small (3 liters per person per day); far less than coal

### @lu2021: pumped hydro for 100% renewables in SE Asia

-   eliminating fossil fuels = triple electricity production
-   electricity demand needs to triple to electrify transport (via electric vehicles), heating (via electric heat pumps and electric furnaces) and chemical synthesis

### @koohi-fayegh2020

-   pumped hydro provides storage from hours to weeks

### @hunt2020: global potential of pumped hydro

-   pumped hydro provides storage from hours to weeks

### @blakers2019: pathway to 100% renewable elctricity

-   deep electrification is key to future, reduce emissions by ¾

### @nadeem2019

-   a range of storage technologies are under development

### @blakers2017: 100% renewables in Aussie

-   examined the amount of storage required to support a 100% renewable electricity system in Australia which derives 90% of its energy from variable wind and solar and 10% from existing hydro and bioenergy sources
-   1 GW (storage) and 20 GWh (energy) per million people
-   spillage of extra (\~ 5-25%) preferable to overbuilding storage

## References

::: {#refs}
:::
